\todonum{acks}
I want to thank my supervisor Ranjit Jhala for all the inspiration. 

I want to thank my committee members 
Samuel R. Buss, 
Cormac Flanagan for inspiring my work,
Sorin Lerner for direction on the PL group, 
and Daniele Micciancio for his strong Haskell enthousiasm 
for being in my committee. 

Next thank the group and specifically my coauthors
Patric Rondon,
Alexander Bakst, 
Eric Seidel for being there. 
and the rest of the group for the supporting and friendly working environment.

I want to thank all these people who hosted me in my internships. 
%
The Opa group in Paris who foresaw that types can help industrial development.
%
Then, Dimitrios Vytiniotis and Simon Peyton-Jones
at Microsoft Research Cambridge
who were the first ghc people who believed in Liquid Haskell
and have been still supporting both the project and me personally.
%
I want to thank Jeff Polakow and all the people at Awake Networks 
for giving me the opportunity to spend an awesome summer in Mountain View 
and use Liquid Haskell in real world development code. 
%
Last but not least, 
I want to thank Daan Leijen 
for hosting me at Microsoft Research Redmond 
and mentoring me since then, giving me all his insightful
knowledge and  about research and life in general.

Finally, I want to thanks all the family and friends, around the world,
who supported me throughout this journey.
%
All the people from San Diego, 
who made me feel San Diego as home, with all the goods and bads. 
%
All the friends I met in the infinite travelling in these five years, 
that made me realize how knowledge unites people from different parts of the globe.
%
All my friends and family in Greece that never let me forget that
whenever I want I have a paradise waiting for me.  
%
Specifically, my parents Paraskeui and Nikos and my sister Marianna
who staid up all these Sunday nights to talk to my and keep me sane. 


It is great that the destination is reached, 
but after all, it is the journey that matters. 




\todonum{paper acks: make sure this starts in new page}
Chapter 2, in full, is a reprint of the material as it appears in
Numerical Grid Generational in Computational Fluid Mechanics~2009.
Smith, Laura; Smith, Jane~D., Pineridge Press,~2009. The dissertation
author was the primary investigator and author of this paper.

Chapter 3, in part, has been submitted for publication of the material
as it may appear in Education Mechanics,~2009, Smith, Laura; Smith,
Jane~D., Trailor Press,~2009. The dissertation author was the primary
investigator and author of this paper.

Chapter 5, in part is currently being prepared for submission for
publication of the material. Smith, Laura; Smith, Jane~D\@. The
dissertation author was the primary investigator and author of this
material.
