\documentclass[10pt]{article}
\usepackage{ucsdcseletterhead}
\usepackage{parskip}
\usepackage[margin=1in]{geometry}
\usepackage{mathptmx}
\usepackage{courier}
\usepackage[scaled=.92]{helvet}
\usepackage[T1]{fontenc}
\pagestyle{empty}
\begin{document}
\hfill\today

Kim Barrett, Dean\\
Graduate Studies\\
402 University Center, 0003\\
UC San Diego\\
La Jolla, CA 92093-0003

\vskip\baselineskip
Dear Dean Barrett:

We request permission for Niki Vazou to use the following publication
in her doctoral dissertation. Ms.\ Vazou was the principle
researcher/author on these papers.

Vazou, Niki; Seidel, Eric L.; Jhala, Ranjit.
``LiquidHaskell: Experience with Refinement Types in the Real World '', Haskell, 2014.

Vazou, Niki; Seidel, Eric L.; Jhala, Ranjit; Vytiniotis, Dimitrios; Peyton-Jones, Simon.
``Refinement Types for Haskell'', ICFP, 2014.

Vazou, Niki; Rondon, Patric M.; Jhala, Ranjit. ``Abstract Refinement Types'', ESOP, 2013.

Vazou, Niki; Bakst, Alexander; Jhala Ranjit.
``Bounded Refinement Types,'' ICFP, 2015.

\vspace{10pt}

Moreover, we request permission for Niki Vazou to use material that has been
submitted for publication. Ms. Vazou was the principal
researcher/author on these papers.

Vazou, Niki; Choudhury, Vikraman; Scott, Ryan G.; Newton, Ryan R.; Jhala, Ranjit.
``Refinement Reflection: Parallel Legacy Languages as Theorem Provers''.

Vazou, Niki; Polakow, Jeff.
``Verified Parallel String Matching in Haskell''.


\vskip3\baselineskip
\parbox[t]{.5\textwidth}{%
Professor Ranjit Jhala\\
Chair of Committee}%
\parbox[t]{.5\textwidth}{%
Niki Vazou}
\end{document}
