\section{Implementation: \toolname}
Here we give some more examples on how we can use \toolname.
We start by proving termination on mutual recursive functions, 
using lexicographical ordering. 
%
Then we describe how we proved functional correctness on 
two commonly used functions, namely \texttt{ByteString}
and \texttt{Text}.

\subsection{Proving Termination}

   Next, consider the Ackermann function.
   %
   \begin{code}
     ack m n 
       | m == 0    = n + 1
       | n == 0    = ack (m-1) 1 
       | otherwise = ack (m-1) (ack m (n-1))
   \end{code}
   %
   There exists no integer termination metric that decreases at each recursive call.
   %
   However @ack@ terminates because at each call \emph{either}
   @m@ decreases \emph{or} @m@ remains the same and @n@ decreases. 
   %
   In other words, the pair @(m,n)@ strictly decreases according to
   \emph{lexicographic} ordering. 
   %
   To capture this requirement we extend termination metric
   from an integer to a list of integers
   and at each recursive call we check that this list is
   lexicographically decreasing.
   %
   In the case of
   @ack@ this list will simply be the parameters @m@
   and @n@:
   %
   \begin{code}
     ack :: m:Nat -> n:Nat -> Nat / [m,n]
   \end{code}
   %
   Thus, \toolname uses lexicographic ordering on 
   a list of natural numbers to prove termination.
   %
   Termination metrics could be generalized to 
   any \emph{well-found} metric.
   
   \spara{Mutual Recursion}
   %
   Equipped with termination metrics
   \toolname instantiates a powerful
   termination checker that like~\citep{XiTerminationLICS01}
   proves termination even for mutual recursive functions.
   %
   Consider the mutual recursive functions @isEven@ and @isOdd@
   \begin{code}
   {-@ isEven :: n:Nat -> Bool / [n, 0] @-}
   {-@ isOdd  :: n:Nat -> Bool / [n, 1] @-}
   
   isEven 0 = True
   isEven n = isOdd $ n-1
   
   isOdd n  = not $ isEven n 
   \end{code}
   Each call terminates as either @isEven@
   calls @isOdd@ with a decreasing argument, 
   or the argument remains the same, and @isOdd@
   calls @isEven@ that should then decrease the argument.
   % 
   We capture this reasoning using two lexicographic pairs:
   each function has its own metric, 
   and when @isEven@ calls @isOdd@
   the metric of the caller $(n, 0)$
   should be greater that callee's metric
   $(n-1, 1)$.
   %
   Similarly, at @isEven@'s call-site 
   \toolname verifies that	
   $(n, 1) > (n, 0)$.
   %
   For example, the call @isEven m@
   will fire the decreasing metric sequence
   $(m, 0) > (m-1, 1) > (m-1, 0) > (m-2, 1) > \dots$
   that ultimate terminates for \textit{any}
   natural number $m$.


\subsection{Bytestring}\label{sec:bytestring}
\input{bytestring}

\subsection{Text}\label{sec:text}
\input{text}
