% \todonum{acks}
I want to thank Ranjit Jhala for being such a great supervisor. 
%
It is his unique way to interpret, simplify, and beautify my 
thoughts
and his strong willingness and enthusiasm to do useful work 
that directed my research and influenced the way I am thinking. 

A big thanks to my committee members 
Sam Buss, 
Cormac Flanagan,
Sorin Lerner, 
and Daniele Micciancio for their interest and insights in my work. 

Next, I want to thank all my labmates in the UCSD programming languages group and specifically my 
collaborators
Eric Seidel, 
Alexander Bakst,
Pat Rondon, and 
Valentin Robert
for providing a supportive, friendly, and inspiring working environment.

I own a big thanks all these people who hosted me in my internships. 
%
The Opa group in Paris who foresaw that types can help industrial development.
%
Dimitrios Vytiniotis and Simon Peyton-Jones
at Microsoft Research Cambridge
who were the first Haskellers who believed in \toolname
and have been still supporting both the project and me personally.
%
I want to thank Jeff Polakow, Gabriel Gonzalez, and all the people at Awake Networks 
for giving me the opportunity to spend an awesome summer in Mountain View 
and use \toolname in real world development code. 
%
A huge thanks to the two main people responsible for my great summer at 
Microsoft Research Redmond.
%
Rustan Leino who is the innovator in automatic software verification
and yet is always open to discuss, in his unique friendly and humble manner, 
how Dafny and \toolname could be further improved and influenced from one another. 
% 
Last but not least, I want to thank my amazing internship host
Daan Leijen for mentoring me during my internship and since then, 
sharing with me all his insightful knowledge and experience 
about research and life in general.

Thanks to all the Haskell community 
for the enthusiasm and support expressed for \toolname 
from its earliest stages. 
%
I am so proud to be a member of this 
supportive, friendly, and respectful
community! 
%
I want to thank those who introduced me to 
Haskell and Computer Science in general, 
Nikolaos Papaspyrou and Stathis Zachos. 


Next, I want to thank all my San Diego friends who were next to me 
during times of both joy and sorrow and 
who turned this beautiful place into 
my second home.
% 
Sincere thanks to all my women friends
Sohini, Minu, Augusta, Margherita, Marta, and Karyn
who kept me sane and happy during my male dominated Ph.D..
%
I want to thank 
Andreas, Dimos, Vasilis, Konstantinos, Andreas
and the rest of the Greeks in San Diego who help me 
keep my culture and roots while so far away from my country.
%
Finally, a great thanks to the crew and friends in Pappalecco, 
the coffee place where most of the \toolname code and papers were written in. 


Last but not least, I want to thank my family and friends, around the world,
who supported me throughout this journey.
%
All the people I met in my numerous trips in these five years, 
that made me realize how knowledge unites people from different parts of the globe.
%
All my friends and family in Greece that despite the distance, 
always welcome me during my escapes in my home country. 
%
Specifically, my parents Voula and Nikos and my sister Marianna
who stayed up all these Sunday nights for our online calls. 

It is great that the destination is reached, 
but after all, it is the journey that matters. 

\begin{flushright}
Niki Vazou \\ December 2016
\end{flushright}

\newpage

Chapter~\ref{chapter:tool} contains material adapted from the following publication:
\noindent N. Vazou, E. Seidel, and R. Jhala,
``LiquidHaskell: Experience with Refinement Types in the Real World'', 
Haskell, 2014.

Chapter~\ref{chapter:refinedhaskell} contains material adapted from the following publication:
\noindent N. Vazou, E. Seidel, R. Jhala, D. Vytiniotis, and S. Peyton-Jones,
``Refinement Types for Haskell'', 
ICFP, 2014.

Chapter~\ref{chapter:abstractrefinements} contains material adapted from the following publication:
\noindent N. Vazou, P. Rondon, and R. Jhala,
``Abstract Refinement Types'', 
ESOP, 2013.

Chapter~\ref{boundedrefinements} contains material adapted from the following publication:
\noindent N. Vazou, A. Bakst, and R. Jhala,
``Bounded Refinement Types'',
ICFP, 2015.


Chapter~\ref{refinementrflection} has been submitted for publication of the material as it may appear in PLDI 2017:
\noindent Vazou, Niki; Choudhury, Vikraman; Scott, Ryan G.; Newton, Ryan R.; Jhala, Ranjit.
``Refinement Reflection: Parallel Legacy Languages as Theorem Provers''.

Chapter~\ref{stringmatcher} has been submitted for publication of the material as it may appear in ESOP 2017:
\noindent Vazou, Niki; Polakow, Jeff.
``Verified Parallel String Matching in Haskell''.

The dissertation author was the primary investigator and author of these papers.